\def\AS#1{{\color{red}#1}}  % print O in red
\def\AS#1{}       % don't print O
\def\red#1{{\color{red}#1}}
\def\green#1{{\color{green}#1}}
\def\redtt{\color{red}\tt}
\def\bluett{\color{blue}\tt}
\def\bottom{\perp}
\font\manual=manfnt
\def\dbend{{\manual\char127}}
\def\eqq{{\buildrel?\over=}}
%\def\B{\red{$\cal B$}}
\def\C{\red{$\cal C$}}
\def\J{\red{$\cal J$}}
\def\I{{\cal I}}
\def\Z{{\bf Z}}
\def\Q{{\bf Q}}
%\def\C{{\bf C}}
\def\N{{\bf N}}
\def\R{{\bf R}}
\def\Oo{{\cal O}}
\def\CL{\mathop{\rm CL}}
\def\const{\mathop{\rm const}}
\def\erf{\mathop{\rm erf}}
\def\res{\mathop{\rm res}\nolimits}
\def\Proj{\mathop{\rm Proj}}
\def\tc{\mathop{\rm tc}}
\def\IDA{{}_{\rm DA}\int}
\def\DDA{D_{\rm DA}}
\def\arctanDA{\arctan_{\rm DA}}
\def\logDA{\log_{\rm DA}}
\def\cDA{constant${}_{\rm DA}$}
\def\fracDDA#1#2{\frac{{\rm d}#1}{{\rm d}_{\rm DA}#2}}
\def\Ied{{}_{\epsilon\delta}\int}
\def\Ded{D_{\epsilon\delta}}
\def\arctaned{\arctan_{\epsilon\delta}}
\def\loged{\log_{\epsilon\delta}}
\def\fracDed#1#2{\frac{{\rm d}#1}{{\rm d}_{\epsilon\delta}#2}}
%\documentclass[notes=hide]{beamer}   % Without note
\documentclass[handout]{beamer}   % handout mode 
%\usepackage[hyphen]{url}
%\usepackage[final]{pdfpages}
\newcommand{\Qxiblock}[2]{Q_{#1} x_{#1,1},\ldots,x_{#1,#2_{#1}}}
%\usepackage{tikz}
\usepackage{hyperref}

%\mode<presentation>

\usetheme{Warsaw}

\setbeamertemplate{navigation symbols}{}

\usepackage{verbatim}
%\usepackage{enumitem}

\usepackage{color}
\bibliographystyle{alpha}
\title{Proving an Execution of an Algorithm Correct?\\
The case of Polynomial Factorisation}
\author{James Davenport\\% \& Tim French\\
\tt masjhd@bath.ac.uk\\
With Edgar Costa (MIT), Alex Best (Imperial), Mario Carneiro (CMU)}
%(Thanks to RJB for the improved title)}
\institute{University of Bath\\Thanks to IPAM at UCLA for prompting this, and many colleagues, especially at Dagstuhl seminar 23401, for input\\Partially supported by EPSRC  grant EP/T015713}%(visiting Waterloo)}
\date{17 November 2023}
\expandafter\def\expandafter\insertshorttitle\expandafter{%
  \insertshorttitle\hfill%
  \insertframenumber\,/\,\inserttotalframenumber}
\begin{document}
\iffalse
\cite{Hoffmannetal2023a} for primality.
\fi
\frame{
\titlepage
}
\begin{frame}[fragile]
\frametitle{Factorisation in $\Z[x]$: the implicit contract}
Consider the case of polynomial factorisation (over the integers). 
We might say that $x^8+5x^4+4$ factors as $(x^4+1)\cdot(x^4+4)$, \pause and this is pretty trivial (given a suitable library!) to verify.
\par\pause
But in fact there’s an implicit contract that the factors returned are themselves irreducible,\pause{} and with that this is no longer the complete factorisation: rather $(x^4+1)*(x^2+2x+2)*(x^2-2x+2)$.
\par\pause
Now how do we verify the accuracy of the factorisation AND the implicit contract?
\par\pause
We will explain how computer algebra systems do this, \pause and look at the progress the speaker and colleagues have made with formalising this.

\end{frame}
\begin{frame}[fragile]
\frametitle{General Situation}
Do I believe the output from my (complicated, optimised, unverified) computer algebra system?
\par\pause
See JHD's paper at CICM 2023 \cite{Davenport2023b}, but note that the same question, in different settings, was asked by Mehlhorn \cite{Mehlhorn2011a} in 1999.  % date from Tobias Nipkow
\par\pause
\cite{Davenport2023b} looked at three examples.
\pause
\begin{description}[<+->]
\item[Polynomial Factorisation]$f=f_1f_2\cdots f_k$ \emph{and} the $f_i$ is irreducible.
\item[Integration]The assertion ``unintegrable'' is correct.
\item[Real Algebraic Geometry]The assertion that the semi-algebraic variety is empty (UNSAT) is correct.
\end{description}
\pause
The last is the most important question for JHD/applications , but factorisation is easy to explain and a good case study in its own right.
\end{frame}
\begin{frame}[fragile]
\frametitle{Polynomial Factorisation}
The base case is polynomials in $\Z[x]$. 
\begin{problem}[Factorisation]\label{prob:fact}
Given $f\in\Z[x]$, write $f=\prod f_i$ where the $f_i$ are \emph{irreducible} elements of $\Z[x]$.
\end{problem}
\pause Verifying that $f=\prod f_i$ is, at least relatively, easy. \pause The hard part is verifying that the $f_i$ are \emph{irreducible}. \pause JHD knows of no implementation of polynomial factorisation that produces any evidence, let alone a proof, of this.
\iffalse
\pause
\begin{problem}[Factorisation in this style]\label{prob:fact-basic}
Given $f\in\Z[x_1,\ldots,x_n]$, produce
        \begin{description}
                \item[either]a proper factor $g$ of $f$,
\item[or]$\bottom$ indicating that no such $g$ exists.
        \end{description}
\end{problem}
\fi
\pause\par
We may as well assume $f$ is square-free \pause(this would be a rather separate verification question).
\end{frame}
\begin{frame}[fragile]
\frametitle{Algorithm}
The basic algorithm goes back to \cite{Zassenhaus1969}: step M is a later addition \cite{Musser1975a}, and the  H' variants are also later.\pause
\begin{enumerate}[<+->]
\item Choose a prime $p$ (not dividing the leading coefficient of $f$) such that $f\pmod p$ is also square-free. Assume $f$ monic for simplicity of this exposition.
\item\label{step:p} Factor $f$ modulo $p$ as $\prod f_i^{(1)} \pmod p$.
\item[M]Take five $p$ and compare the factorisations.
\item If $f$ can be shown to be irreducible from modulo $p$ factorisations, return $f$.
\item Let $B$ be such that any factor of $f$ has coefficients less than $B$ in magnitude, and $n$ such that $p^n\ge 2B$. [Landau--Mignotte]
\item Use Hensel's Lemma to lift the factorisation to $f=\prod f_i^{(n)} \pmod {p^n}$
\item[H]\label{step:H} Starting with singletons and working up, take subsets of the $f_i^{(n)}$, multiply them together and check whether, regarded as polynomials over $\Z$ with coefficients in $[-B,B]$, they {\color{red}divide} $f$ --- if they do, declare that they are irreducible factors of $f$.
\end{enumerate}
\end{frame}
\begin{frame}[fragile]
\frametitle{Algorithm Notes (1)}
\begin{enumerate}[<+->]\item[H']\label{step:H'}Use some alternative technique, originally \cite{Lenstraetal1982}, but now e.g. \cite{Abbottetal2000a,Hartetal2011a} to find the true factor corresponding to $f_1^{(n)}$, remove $f_1^{(n)}$ and the other $f_i^{(n)}$ corresponding to this factor, and repeat.
\item[\dbend]In practice, there are a lot of optimisations, which would greatly complicate a proof of correctness of an implementation of this algorithm.
\end{enumerate}
\pause
\begin{quote}
We found that, although the Hensel construction is basically neat and simple in theory,
the fully optimised version we finally used was as nasty a piece of code to write and
debug as any we have come across \cite{MooreNorman1981}.
\end{quote}
\pause
There are many optimisations around {\color{red}divide} that probably {\color{red}ought} to be passed to the theorem prover. \pause We have a product modulo $p^n$, which we can regard as over $\Z$ with coefficients in $[-p^n/2,p^n/2]$, but this might be outside $[-B,B]$. \pause Also, if $g|f$, $\tc(g)|\tc(f)$.
\end{frame}
\begin{frame}[fragile]
\frametitle{Algorithm Notes (2)}
Since if $f$ is irreducible modulo $p$, it is irreducible over the integers, the factors produced from singletons in step \ref{step:H} are easily proved to be irreducible.  Unfortunately, the chance that an irreducible polynomial of degree $n$ is irreducible modulo $p$ is $1/n$.
\par\pause
So the proof that, say, a factor over $\Z$ produced from multiplying four factors modulo $p^n$, is that none of the individual ones, and none of the pairs, gave rise to a true divisor.
\pause
Note that we don't need to check the triples \emph{in the proof} because if the triple were a factor, since the quadruple is a factor, the missing singleton would also be a factor.
\end{frame}
\begin{frame}[fragile]
\frametitle{Algorithm Notes (3)}
%\par\pause
A factorisation algorithm could, even though no known implementation does, relatively easily produce the required information for a proof of irreducibility unless the recombination step is required. \pause
\begin{enumerate}[<+->]
\item[Note]that \emph{verifying} the Hensel lifting, the ``nasty piece'' from \cite{MooreNorman1981} is easy: the factors just have to have the right degrees from the factorisation of $f \pmod p$ and multiply to give $f\pmod{p^n}$.
\item[\dbend]Building test cases for the various edge cases was extremely difficult.
\end{enumerate}
\pause
Step [H] is relatively easy to verify: this combination divides and no relevant smaller combination divides. \pause The variants in [H'] are challenging: I have not found an easy route. \par\pause
If [H'] finds a factor that is a product of $k$ $p$-adic factors, then we can use [H] to verify this by checking that the $2^{k-1}-1$ ``small'' subsets do not give factors.
\par\pause
But if [H'] says ``irreducible'', I know no easy proof.
\end{frame}
\begin{frame}[fragile]
\frametitle{Progress at Dagstuhl (October 2023)}
\pause
\begin{enumerate}[<+->]
\item
We can extract from the implementation in FLINT \cite{Flint2023a} of the algorithm with [H], at essentially no cost, the key data that we believe a verifier would need to confirm the irreducibility.
\item But this is not necessarily the most efficient verification.
\item We \emph{think} that a more efficient verification would need negligibly more work.
\item We haven't built a verification.
\item The ``hard'' theorems, e.g. $M$, are (being) stated (LEAN), but what about the ``easy'' ones, mappings such as ``regarded as polynomials over $\Z$ with coefficients in $[-B,B]$''?
\item Needs more theorem prover input.
\item[But]We have discovered improvements to FLINT, and at least one new research question in computer algebra.
\item[+]FLINT also has [H'], but we haven't looked at this yet.
\end{enumerate}
\end{frame}
\begin{frame}[fragile]
\frametitle{So what is the certificate?}
\pause
\begin{enumerate}[<+->]
\item The 5(?) Musser primes (or the useful subset)
\item The factorisations modulo these
\item[*]Need to verify that these are irreducible.
\item The chosen $p$ and $n$ (probably $B$)
\item[*]Need to verify that $p^n>2B$: need a library proof of $B$.
\item The set $S$ of factors modulo $p^n$
\item[*]Need to check they match the mod $p$ ones, and multiply. We have already proved irreducibility of the mod $p$ versions
\item The partition $S=\bigcup S_i$ that corresponds to the true factorisation.
\end{enumerate}
\pause
Thanks to Tobias Nipkow for asking this explicitly.
\end{frame}
\iffalse
\begin{frame}[fragile]
\frametitle{Further Reflections}
\begin{enumerate}
\item[M]Take five $p$ and compare the factorisations.
\end{enumerate}
\pause
Not just ``take the best''. \pause Rather we look for incompatibilities, so if a degree 4 factors as 3,1 modulo one prime and 2,2 modulo another, it's actually irreducible, and so on.
\par\pause
\cite{Musser1975a} suggests taking five primes, though more recently \cite{LuczakPyber1997} show that, if the Galois group is $S_n$, seven is asymptotically right.
\par\pause
For any degree $d$, the probability that a random polynomial with coefficients $\le H$ has Galois group $S_n$ tends to 1 as $H$ tends to infinity. \pause \cite{DavenportSmith2000} looks at other Galois groups.
\end{frame}
\fi

\begin{frame}[allowframebreaks]
\frametitle{Bibliography}
\bibliography{../../../../../jhd}
\end{frame}

\end{document}
\begin{frame}[fragile]
\frametitle{}
\pause
\begin{itemize}[<+->]
\item 
\end{itemize}
\end{frame}
\pause
	\begin{description}
		\item[]
	\end{description}
\begin{frame}[fragile]
\frametitle{}
\pause
\begin{enumerate}[<+->]
\item 
\item 
\item 
\item 
\item 
\end{enumerate}
\end{frame}
\begin{frame}
\begin{frame}
